Using Occupancy Theory and Hierarchical Models to Estimate Otter Population in Southeastern Minnesota

### 2010-02-27 Paper.tex
Edited Occupancy and Independence sections

    The main problem with these methods of sampling data is that it is possible
    for an animal to occupy a site, but not be detected by the observer. This
    means that the site is recorded as unoccupied, but there really is an animal
    there. This is called a false negative. Marking an animal present when there
    really is no animal is also a possible problem, but is thought to be much
    less likely and is often ignored. We will discuss this issue more below.

    \subsection{Assumptions}
    There are some assumptions that are made in most models of occupancy.
    \cite{MacKenzie2006}. Without these assumptions, the models become much more
    complicated. One common assumption is no false positives, which means if a
    site is recorded as occupied we say it must actually be occupied. A second
    common assumption is that presence and detection probabilities are constant
    across sites and surveys. This simplifies the estimation of each parameter.
    A third common assumption is that there is independence of detection between
    sites. A fourth common assumption is that the surveys are being conducted on
    a closed population. This means that no animals leave the area or join the
    area during the time of the surveys.

    Obviously, not all of these assumptions are always true, but they are often
    necessary in order to get a model that is simple enough to use. There often
    are possible ways to get rid of these assumptions, but it complicates the
    model. In this paper, we will only talk about attempting to get rid of the
    assumption of no false positives, and the assumption of independence of
    parameters across sites.

    \subsection{Types of occupancy models}
    All of the models discussed in this paper are single-species, single-season
    models. It is possible to modify all of these models to be used for multiple
    seasons or multiple species, but it would make them more complicated.

    The most simple way to measure occupancy is to simply divide the number of
    sites where an animal was found to be present by the total number of sites.
    This is called na\"ive occupancy \cite{MacKenzie2006}. It is generally
    accepted that this estimate is too low because animals are not detected at
    all sites where they are present, which is the issue of false negatives
    mentioned above. Other occupancy models attempt to do better than this
    simple estimate.

    One way to create more accurate models is to use a Hierarchical model as a
    way to think about the parameters that affect occupancy. This allows us to
    take into account that when an animal is not detected, it could either not
    be there, or it could be there and just was not detected. This could either
    be due to a false negative, where the observer did not see the indication --
    for example they did not see the animal, track, etc. -- or it could be due
    to the fact that the animal does occupy the site but there was no indication
    of the animal at the site -- for example, no track, no vocalization, etc.

@@

    \subsection{False positive rates}
    The lack of consistency among the flights is strong evidence that the data
    contain false positives where scientists think they see a track when it does
    not exist. Not accounting for such a process can lead to major biases; as
    the number of flights increases, the occupancy rate estimate will approach
    1 \cite{Royle2006}. Unlike most occupancy models where scientists assume 
    that detection of the species in question guarantees its presence \cite
    {MacKenzie2006}, we introduced a false detection process whereby an observer 
    can mark down the existence of a track when in reality there is none. This 
    false detection process can occur as long as there is no track, whether or 
    not the site is occupied.
    \subsection{Testing the independence assumption}
    \label{shat}
    First we considered using Markov Chains to come up with a long term
    probability of detection, which we called \(r\). We let \(p\) be the number
    of \(0\)'s followed by a \(0\) as a fraction of all successive pairs that
    start with a \(0\). We let \(q\) be the number of \(1\)'s followed by a
    \(1\) as a fraction of all successive pairs that start with a \(1\). These
    probabilities can be represented by the matrix:
    \[M=\begin{bmatrix}
        p & 1-p \\
        1-q & q
    \end{bmatrix}\text{.}\]

    The long term probability of detection can be found from this by finding
    \(\lim_{n \to \inf}M^n\), which returns
    \[\begin{bmatrix}
        1-r & r \\
        1-r & r
    \end{bmatrix}\text{,}\]
    where \(r\) is the probability of a \(1\), which is the long term
    probability of detection.

    We calculated \(p\), \(q\), and \(r\) for every flight in our data, using
    \(n = 500\). We then realized that \(r\) is actually not so useful in
    analyzing our data. It does not take into account whether or not a site is
    occupied, or whether or not a track was laid down on that site, or any of
    the ideas behind the hierarchical model. These are the parameters that we
    eventually want to find out something about, and the value of \(r\) cannot
    help us get there. The idea to use Markov Chains was still useful, however.

    Although we rejected the idea of using \(r\) as a way to analyze the data,
    we realized that \(p\) and \(q\) could be used to test the viability of the
    independence assumption. If there is independence between sites, the
    difference of the proportions \(p\) and \(1-q\) should be approximately
    \(0\). This would mean that the probability that a \(0\) follows a \(0\) is
    about the same as the probability that a \(1\) follows a \(0\), meaning that
    it does not depend on what the observation is at the previous site. Looking
    at boxplots of \(p-(1-q)\), we see that data from flights that record
    alternating sites may be more independent than data from flights that record
    all sites, which suggests a possible benefit to using the alternating
    sampling method. The boxplots for the Whitewater and Zumbro Rivers looked
    similar to the Mississippi:

    \begin{center}
        \includegraphics[scale=0.55]{PQBoxPlots}
    \end{center}

    We found no obvious trend indicating one site size to be more independent
    than the others.

    While this is an interesting observation, we would like to be able to
    quantify it better by finding the distribution of our statistic. This would
    allow us to make statements about how independent the data is, and if the
    difference in independence between different types of data is significant.
    This statistic seems hard to get, but we considered a modification. Let
    \(S\) be the number of successive \(0\), \(1\)'s minus the number of
    successive \(1\), \(1\)'s per flight. Under the null hypothesis that each
    site is an independent Bernoulli trial with probability \(p\) we were able
    to compute the mean and standard deviation of \(S\), which we used to
    calculate a Z-score. In order to compute these values, we needed to find the
    expected value and variance of each term in \(S\). We let \(A = \text{\# of
    0-1's}\), \(B=\text{\# of 1-1's}\) and \(p=P(``1")\text{ in the sequence}\).

    In order to find \(E[\hat{s}]\), we first needed to find \(E[A]\) and
    \(E[B]\).

    % E[s-hat]
    \begin{equation*}
        \begin{split}
        E[\hat{s}]& =E[A]-E[B] \\
                  & =(n-1)\left((1-p)p-p^2\right)
        \end{split}
    \end{equation*}

    % Var[s-hat]
    \begin{equation*}
        \begin{split}
        Var[\hat{s}]& =Var[A-B] \\
                    & =Var[A]+Var[B]-2Cov[A,B] \\
                    & =(1-p)p(-1+n+8p-4np-20p^2+12np^2)
        \end{split}
    \end{equation*}

    Simulations of \(S\) show that the distribution of \(S\) is roughly normal.
    So using our calculations for the mean and the standard deviation, we
    computed a Z-score for each flight, where we use a plug-in estimate for
    \(p\) to be the proportion of \(1\)'s per flight. We said that Z-scores with
    an absolute value greater than \(2\) suggest that the assumption of
    independence between sites is violated for that flight. We found that, in
    general, alternating sites are more "independent" than all sites as measured
    by this statistic. However, alternating sites do not guarantee independence.
    For instance, six of eleven flights along the Zumbro River during 2003 still
    have significant Z-scores with alternating sites. There were no noticeable
    trends for this statistic with respect to observer, days since snow, snow
    event, or site size.

@@

>>>>

    The main problem with these methods of sampling data is that it is possible
    for an animal to occupy a site, but not be detected by the observer. This
    means that the site is recorded as unoccupied, but there really is an animal
    there. This is called a false negative. Marking an animal present when there
    really is no animal is also a possible problem, but is thought to be much
    less likely and is often ignored. We will discuss this issue more in later sections.

    \subsection{Assumptions}
    As is true for all modeling, there are some assumptions made in most models of occupancy 
in order to create simple enough models \cite{MacKenzie2006}. One common assumption is no false positives, which means if a
    site is recorded as occupied we say it must actually be occupied. A second
    common assumption is that presence and detection probabilities are constant
    across sites and surveys. This simplifies the estimation of each parameter.
    A third common assumption is that there is independence of detection between
    sites. A fourth common assumption is that the surveys are being conducted on
    a closed population. This means that no animals leave the area or joins the
    area during the time of the surveys.

    There are different methods for dealing with violations of these assumptions, some 
more developed than others.  In this paper, we will only talk about attempting to get rid of the
    assumption of no false positives and the assumption of independence of
    parameters across sites.

    \subsection{Types of occupancy models}
    All of the models discussed in this paper are single-species, single-season
    models. It is possible to modify all of these models to be used for multiple
    seasons or multiple species, at the cost of added complexity.

    The most simple way to measure occupancy is to simply divide the number of
    sites where an animal was found to be present by the total number of sites.
    This is called na\"ive occupancy \cite{MacKenzie2006}. It is generally
    accepted that this estimate is too low because animals are not detected at
    all sites where they are present, due to the issue of false negatives
    mentioned above. Other occupancy models attempt to do better than this
    simple estimate.

    One way to create a more accurate model is to attempt to model the biological 
proccess that is occuring.  In order to do this, we can use a Hierarchical model as a
    way to think about the parameters that affect occupancy. This allows us to
    take into account that when an animal is not detected, it could either not
    be there, or it could be there and just was not detected. This could either
    be due to a false negative, where the observer did not see the indication --
    for example they did not see the animal, track, etc. -- or it could be due
    to the fact that the animal does occupy the site but there was no indication
    of the animal at the site -- for example, no track, no vocalization, etc.

@@

    \subsection{False positive rates}
    The lack of consistency among the flights is strong evidence that the data
    contain false positives where scientists think they see a track when it does
    not exist. Not accounting for such a process can lead to major biases; as
    the number of flights increases, the occupancy rate estimate will approach
    1 \cite{Royle2006}. Unlike most occupancy models where scientists assume 
    that detection of the species in question guarantees its presence \cite
    {MacKenzie2006}, we introduced a false detection process whereby an observer 
    can mark down the existence of a track when in reality there is none. This 
    false detection process can occur as long as there is no track, whether or 
    not the site is occupied.

    \subsection{Testing the independence assumption}
    \label{shat}
    Since one of the assumptions of occupancy modeling is that there is 
	independence between sites, we wanted to see if this assumption held true 
	for our data.  In order to do this, we calculated the statistic \(S\), which
	we define as the number of successive \(0\), \(1\)'s minus the number of
    successive \(1\), \(1\)'s per flight. Under the null hypothesis that each
    site is an independent Bernoulli trial with probability of a \(1 = p\) we were able
    to compute the mean and standard deviation of \(S\), which we used to
    calculate a Z-score, where \(Z = (S – E[S])/SD[S]\). In order to compute these values, we needed to find the
    expected value and variance of each term in \(S\). These caclulations can be
	found in the appendix.

    Simulations of \(S\) show that the distribution of \(S\) is roughly normal.
    So using our calculations for the mean and the standard deviation, we
    computed a Z-score for each flight, where we use a plug-in estimate for
    \(p\) to be the proportion of \(1\)'s per flight. We said that \(|Z| > 2\) suggests that the assumption of
    independence between sites is violated for that flight. We found that, in
    general, alternating sites are more "independent" than all sites as measured
    by this statistic. However, alternating sites do not guarantee independence.
    For instance, six of eleven flights along the Zumbro River during 2003 still
    have significant Z-scores with alternating sites. It also appears that there 
	may be slightly more independence when larger site sizes are used, although it is
not clear if this trend is significant.  There 
	were no noticable trends for this statistic with respect to observer, days 
	since snow, or snow event.

@@

%Needs to be fixed/formatted better... I didn't know how to make an appendix
\appendix
	We let \(A = \text{\# of
    0-1's}\), \(B=\text{\# of 1-1's}\) and \(p=P(``1")\text{ in the sequence}\).
    In order to find \(E[s]\), we first needed to find \(E[A]\) and
    \(E[B]\).

    % E[s]
    \begin{equation*}
        \begin{split}
        E[s]& =E[A]-E[B] \\
                  & =(n-1)\left((1-p)p-p^2\right)
        \end{split}
    \end{equation*}

    % Var[s]
    \begin{equation*}
        \begin{split}
        Var[s]& =Var[A-B] \\
                    & =Var[A]+Var[B]-2Cov[A,B] \\
                    & =(1-p)p(-1+n+8p-4np-20p^2+12np^2)
        \end{split}
    \end{equation*}
